# Data sets {#sec-chp11}

```{r}
#| message: false
library(sf)
library(dplyr)
```

## Greater Machester land use data

### Availability {.unnumbered}

The dataset is stored on a gpkg file that can be found, within the structure of this project, under:

```{r}
st_LSOA <- st_read("./data/geodemographics/manchester_land_cover_2011.gpkg")
```

### Variables {.unnumbered}

The variables included in this dataset follow the land use classification of the CORINE Land Cover dataset.

### Source & Pre-processing {.unnumbered}

The data was sourced from [What do 'left behind' areas look like over time?](https://github.com/GDSL-UL/APPG-LBA/blob/main/README.md) and cleaned on Python.

## British administrative boundaries (LSOAs and LAs)

### Availability {.unnumbered}

The dataset for the boundaries of the lower-layer super-output areas (LSOAs) within London is stored as a shapefile that can be found under:

```{r}
st_LSOA <- st_read("data/geodemographics-old/LSOA_2011_London_gen_MHW/LSOA_2011_London_gen_MHW.shp")
```

The dataset for the boundaries of the local authority distrits (LADs) for the UK is stored as a shapefile that can be found under:

```{r}
LA_UK <- st_read("./data/networks/Local_Authority_Districts_(December_2022)_Boundaries_UK_BFC/LAD_DEC_2022_UK_BFC.shp")
```

### Variables {.unnumbered}

For each of the 4,835 LSOAs, the following characteristics are available:

```{r}
names(st_LSOA)
```

where:

-   `LSOA11CD`: Lower-Layer Super-Output Area code
-   `LSOA11NM`: Lower-Layer Super-Output Area code
-   `MSOA11CD`: Medium-Layer Super-Output Area code
-   `MSOA11NM`: Medium-Layer Super-Output Area code
-   `LAD11CD`: Local Authority District code
-   `LAD11NM`: Local Authority District name
-   `RGN11CD`: Region code
-   `RGN11NM`: Region name
-   `USUALRES`: Usual residents
-   `HHOLDRES`: Household residents
-   `COMESTRES`: Communal Establishment residents
-   `POPDEN`: Population density
-   `HHOLDS`: Number of households
-   `AVHHOLDSZ`: Average household size
-   `geometry`: Polygon of LSOA

For each of the 374 LADs, the following characteristics are available:

```{r}
names(LA_UK)
```

where:

-   `OBJECTID`: object identifier
-   `LAD22CD`: Local Authority District code
-   `LAD22NM`: Local Authority District name
-   `BNG_E`: Location Easting
-   `BNG_N`: Location Northing
-   `LONG`: Location Longitude
-   `LAT`: Location Latitude
-   `GlobalID`: Global Identifier
-   `SHAPE_Leng`: Boundary length
-   `SHAPE_Area`: Area within boundary
-   `geometry`: Polygon of LAD

### Projection {.unnumbered}

The shapes of each LSOA are stored as polygons an expressed in the OSGB36 projection:

```{r}
st_crs(st_LSOA)
```

Similarly, the shapes of each LAD are stored as polygons an expressed in the OSGB36 projection:

```{r}
st_crs(LA_UK)
```

### Source & Pre-processing {.unnumbered}

The boundaries for the LSOAs within London can be found directly from the [London Datastore](https://data.london.gov.uk/dataset/statistical-gis-boundary-files-london) website.

The boundaries for the LADs for the UK can be found on the ONS Open Geography Portal [website](https://geoportal.statistics.gov.uk/search?collection=Dataset&sort=name&tags=all(BDY_LAD%2CDEC_2022)). To filter for the London LADs, i.e. the London boroughs, we run the following line of code:

```{r}
LND_boroughs <- LA_UK %>% filter(grepl('E09', LAD22CD)) 
```

## Worldpop population count data for Ukraine

## Census population count data for UK

## Ukraine's administrative boundaries

## Internal migration flows between US metropolitan areas and between London boroughs

### Availability {.unnumbered}

The dataset for the migration flows between US metropolitan areas can be found as a csv file under:

```{r}
df_metro <- read.csv("./data/networks/metro_to_metro_2015_2019_US_migration.csv")
```

The dataset for the migration flows between London boroughs can be found as a csv file under:

```{r}
df_borough <- read.csv("./data/networks/LA_to_LA_2019_London_clean.csv")
```

### Variables {.unnumbered}

For each of the 52,930 movements recorded on the dataset for the migration flows between US metropolitan areas, the following fields are available:

```{r}
names(df_metro)
```

All the fields that start with `MSA_Current_` or `MSA_Previous_` refer to the characteristics of the origin and destination metropolican areas. The relevant fields for the analysis in this book are:

-   `Movers_Metro_to_Metro_Flow_Estimate`: Estimate of number of people moving between origin and destination
-   `Movers_Metro_to_Metro_Flow_MOE`: Margin of error for the above estimate

More details on the methodology to obtain the estimates and the margin of error for each population movement can be found on the [US Census Bureau website](https://www.census.gov/programs-surveys/acs/methodology.html).

For each of the 1,053 movements recorded on the dataset for the migration flows between London boroughs, the following fields are available:

```{r}
names(df_borough)
```

where:

-   `OutLA` is the code corresponding to the origin borough
-   `InLA` is the code corresponding to the destination borough
-   `Moves` is the number of internal migration moves within each flow. Note that the numbers are not integers. This is because of the various scaling processes used to produce the dataset, which are described in more detail in the latest methodology document, which can be found [here](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/migrationwithintheuk/datasets/internalmigrationbyoriginanddestinationlocalauthoritiessexandsingleyearofagedetailedestimatesdataset).

### Source & pre-processing {.unnumbered}

The dataset for the migration flows between US metropolitan areas can be downloaded from the [US Census Bureau website](https://www.census.gov/data/tables/2019/demo/geographic-mobility/metro-to-metro-migration.html). The data was cleaned on Microsoft Excel.

The dataset for the migration flows between London boroughs can be downloaded from the [ONS website](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/migrationwithintheuk/datasets/internalmigrationbyoriginanddestinationlocalauthoritiessexandsingleyearofagedetailedestimatesdataset). The data was cleaned on Microsoft Excel.

## Twitter data on public opinion originated in the US and in the UK

## Reddit data

### Availability

Reddit is a social news website where you can find posts about almost anything. Reddit also has a plug-n-play R package which makes is very easy to get Reddit data via API. Everyone can source their own Reddit data. The data used on fertility rates during the pandemic can be accessed through the following code:

```{r}
comments <- read.csv("data/topic-modelling/comments.csv", header = TRUE)

```

### Variables

For each of the 78,065 comments from Reddit, the following fields are available:

```{r}
names(comments)
```

### Source & Pre-processing

More information on Reddit data and the Reedit API can be found [here](https://rpubs.com/mswofford/redditAPI). Some considerable pre-processing needs to happen to create a Corpus style object using `tidytext` as shown in [Chapter 7](https://www.population-science.net/topic-modelling). Sme insights on using Reddit data for research [here](https://journals.sagepub.com/doi/10.1177/20563051211019004).

## Google mobility data for Italy and the UK

### Availability

[COVID-19 community mobility reports](https://www.google.com/covid19/mobility/) area all freely accessible online.

To access mobility reports for Italy directly, you can the load the following:

```{r}
mobility_it <- read.csv("data/longitudinal-1/2022_IT_Region_Mobility_Report.csv", header = TRUE)
```

To access mobility reports for the UK directly, you can the load the following:

```{r}
mobility_gb <- read.csv("data/longitudinal-1/2021_GB_Region_Mobility_Report.csv", header = TRUE)
```

The datasets for the geographic boundaries can be load the following for Italy:

```{r}
italy_iso3166 <- st_read("data/longitudinal-1/italy_projected_simplified.geojson")
```
With the following CRS: 

```{r}
st_crs(italy_iso3166)
```

and this for the UK

```{r}
uk_iso3166 <- st_read("data/longitudinal-1/uk_projected_simplified.geojson")
```

With the following CRS: 

```{r}
st_crs(uk_iso3166)
```

### Variables

The following variables are available.

```{r}
names(mobility_it)
```

-   **Grocery & pharmacy** - Mobility trends for places like grocery markets, food warehouses, farmers markets, specialty food shops, drug stores, and pharmacies.
-   **Parks Mobility** -  trends for places like local parks, national parks, public beaches, marinas, dog parks, plazas, and public gardens.
-   **Transit stations** -  Mobility trends for places like public transport hubs such as subway, bus, and train stations.
- **Retail & recreation** -  Mobility trends for places like restaurants, cafes, shopping centers, theme parks, museums, libraries, and movie theaters.
- **Residential Mobility** -  trends for places of residence.
- **Workplaces Mobility** - trends for places of work. 

## COVID-19 cases data for London and Rome

### Availability

COVID-19 case data can be downloaded for the UK through the [corovirus data portal](https://coronavirus.data.gov.uk/details/download).

Data for London was pre-prepared for the book and can be accessed here:
```{r}
covid_cases_london <- read.csv("data/longitudinal-2/covid_cases_london.csv", header = TRUE)
```

COVID-19 case data can be downloaded for the Italy through the 
Presidenza del Consiglio dei Ministri - Dipartimento della Protezione Civile [github](https://github.com/pcm-dpc/COVID-19/tree/master/dati-province).

Data for the Lazio region was pre-prepared for the book and can be accessed here:
```{r}
covid_cases_lazio <- read.csv("data/longitudinal-2/covid_cases_lazio.csv", header = TRUE)
```


### Variables

For each of the 34,229 observations in the London datasets, the following characteristics are available.  

```{r}
names(covid_cases_london)
```
We consider `cumCasesBySpecimenDate` the cumulative cases equivalent to the total cases present in the Italian data.

For each of the 15,793 observations in the Lazio datasets, the following characteristics are available. 

```{r}
names(covid_cases_lazio)
```
- **data** - Date and time
- **stato** - Country
- **codice_regione** - Regional Code
- **denominazione_regione** - Region
- **codice_provincia** - Provincial Code
- **denominazione_provincia** - Province (sub-region)
- **totale_casi** - Total number of cases

## Census MSOA data for England and Wales

### Availability {.unnumbered}

The dataset for the demographic census data of each MSOA in England and Wales can be loaded as a csv file from:

```{r}
df_MSOA <- read.csv("./data/machine-learning/census2021-msoa-income.csv")
```

A very similar dataset for the demographic census data of each MSOA in England and Wales which also contains data on the median house price can be loaded as a csv file from:

```{r}
df_housing <- read.csv("./data/machine-learning/census2021-msoa-houseprice.csv")
```

### Variables {.unnumbered}

For each of the 7,080 MSOAs recorded in England and Wales, the following fields are available:

```{r}
names(df_MSOA)
```

For a description of the variables in the columns of df_MSOA, we can load a dictionary for these variables:

```{r}
df_dictionary <- read.csv("./data/machine-learning/Dictionary.csv")
head(df_dictionary)
```

### Source & pre-processing {.unnumbered}

Data on the the census characteristics for different MSOAs can be downloaded from the [Nomis website](https://www.nomisweb.co.uk/census/2021/bulk). Data on the average net household income can be obtained from the [ONS website](https://www.ons.gov.uk/employmentandlabourmarket/peopleinwork/earningsandworkinghours/datasets/smallareaincomeestimatesformiddlelayersuperoutputareasenglandandwales).

Data on the median houseprice for different MSOAs can be downloaded from the [ONS website](https://www.ons.gov.uk/peoplepopulationandcommunity/housing/datasets/hpssadataset2medianhousepricebymsoaquarterlyrollingyear).

All the data has been pre-processed on Microsoft Excel.
