# Data sets {#sec-chp11}

```{r}
library(sf)
```

## Greater London Authority LSOA Atlas

### Availability {.unnumbered}

The dataset is stored on a csv file that can be found, within the structure of this project, under:

```{r}
df_lsoa_atlas <- read.csv("data/geodemographics/lsoa-data-clean.csv")
```

### Variables {.unnumbered}

A description of all the variables included in this dataset can be found in a dictionary csv file, which can be imported as a data frame from:

```{r}
df_lsoa_atlas_dictionary <- read.csv("data/geodemographics/Dictionary-lsoa-data-clean.csv")
```

### Source & Pre-processing {.unnumbered}

The data was sourced from [London Datastore's LSOA Atlas](https://data.london.gov.uk/dataset/lsoa-atlas) and cleaned on Microsoft Excel.

## England and Wales administrative boundaries (LSOAs, MSOAs and LAs)

## Worldpop population count data for Ukraine

## Census population count data for UK

## Ukraine's administrative boundaries

## Internal migration flows between US metropolitan areas and between London boroughs

## Twitter data on public opinion originated in the US and in the UK

## Reddit data

## Google mobility data for Italy and the UK

## COVID-19 cases data for London and Rome

## Census MSOA data for England and Wales

