# Geodemographics {#sec-chp3}

## Dependencies {#sec-sec31}

```{r}
library(sf)
library(stringr)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(viridis)
library(plotrix)
```

```{r}
getwd() #Obtain the working directory, where we will save the d
```

## Data {#sec-sec32}

### Demographic data for Greater London Authority

Data will be from LSOA atlas from GLA. For the session GDC based on some features. For assignment, GDC based on a different set of features. Need to prepare two different datasets.

### Import the data

```{r}

# Import LSOA demographic data for GLA
# Data obtained from link below, but it has been cleaned by Carmen Cabrera-Arnau
# https://data.london.gov.uk/dataset/lsoa-atlas
df_LSOA <- read.csv("./data/geodemographics/lsoa-data-clean.csv")

# Import LSOA boundaries for GLA
# Data obtained from link below
# https://data.london.gov.uk/dataset/statistical-gis-boundary-files-london
st_LSOA <- st_read("./data/geodemographics/LSOA_2011_London_gen_MHW/LSOA_2011_London_gen_MHW.shp")

```

## Geodemographic classification {#sec-sec33}

## Preparing the data for GDC {#sec-sec34}

### Choice of geographic units

```{r}
plot(st_geometry(st_LSOA), border=adjustcolor("gray20", alpha.f=0.4), lwd=0.6)
```
