# Sequence Analysis {#sec-chp4}

This chapter illustrates the use of sequence analysis and [WorldPop](https://www.worldpop.org) raster data to identify trajectories of population decline in Ukraine. Developed in Biology for the analysis of DNA sequencing, sequence analysis offers a novel approach to generate a more holistic understanding of population decline trajectories capturing differences in the ordering, frequency, timing and occurrence of population decline. The *longitudinal* and *categorical* nature is a key feature of the data that can be analysed using sequence analysis. In this chapter, We first show how to manipulate gridded, raster population to create a spatial data frame, and explore national and sub-national population trends and spatial structure of population change. We describe the process of implementing sequence analysis to identify trajectories of population decline in a four-stage process. We first define different levels of population change. Second, we apply measure the difference or similarity between individual area-level trajectories. Third, we use an unsupervised machine learning clustering algorithm to identify representative trajectories. Fourth, we use different visualisation tools to understand key features encoded in trajectories.

The chapter is based on the following references:

@gabadinho2011 describe the functionalities of the `TraMineR` package to visualise and analysis categorical sequence data;

@newsham2022a, @gonzález-leonardo2023 provide empirical applications to identify and study trajectories of population decline across Europe and in Spain;

@tatem2017 provide an overview of the *WorldPop* data project;

@patias2021, @patias2021a provide good examples of the use of sequence analysis to define trajectories of inequality and urban development.

## Dependencies

We use the following libraries:

```{r}
#| include = FALSE
# clear memory and adds tufte templates for plots and maps
rm(list=ls())
source("data-viz-themes.R")
```

```{r}
#| warning: false

# data manipulation
library(tidyverse)
# spatial data manipulation
library(stars)
library(sf)
# download world pop data
library(wpgpDownloadR) # you may need to install this package running `install.packages("devtools")` `devtools::install_github("wpgp/wpgpDownloadR")'
# data visualisation
library(viridis)
library(RColorBrewer)
library(patchwork)
library(ggseqplot) # may need to install by running `devtools::install_github("maraab23/ggseqplot")`
# sequence analysis
library(TraMineR)
# cluster analysis
library(cluster)
```

Key packages to this chapter are `TraMineR`,`stars` and `ggseqplot`. `TraMineR` is the go-to package in social sciences for exploring, analysing and rendering sequences based on categorical data. `stars` is designed to handle spatio-temporal data in the form of dense arrays, with space and time as dimensions. `stars` provides classes and methods for reading, manipulating, plotting and writing data cubes. It is a very powerful package. It interacts nicely with `sf` and is suggested to be superior to `raster` and `terra`, which are also known for their capacity to work with multilayer rasters. `stars` is suggested to [deal with more complex data types](https://r-spatial.github.io/stars/) and [be faster](https://www.r-bloggers.com/2021/05/a-comparison-of-terra-and-stars-packages/) than `raster` and `terra`. `ggseqplot` provides functionality to visualise categorical sequence data based on `ggplot` capabilities. This differs from `TraMineR` which is based on the base function `plot`. We prefer `ggseqplot` for the wide usage of `ggplot` as a data visualisation tool in R.

## Data

The key aim of this chapter is to define representative trajectories of population decline using sequence analysis and WorldPop data. We use WorldPop data for the period extending from 2000 to 2020. WorldPop offers open access gridded population estimates at a high spatial resolution for all countries in the world. WoldPop produces these gridded datasets using a top-down (i.e. dissagregating administrative area counts into smaller grid cells) or bottom-up (i.e. interpolating data from counts from sample locations into grid cells) approach. You can learn about about these approaches and the data available from [WorldPop](https://www.worldpop.org).

WorldPop population data are available in various formats: \* Two spatial resolutions: 100m and 1km; \* Constrained and unconstrained counts to built settlement areas; \* Adjusted or unadjusted to United Nations' (UN) national population counts; \* Two formats i.e. `tiff` and `csv` formats.

We use annual 1km gridded, UN adjusted, unconstrained population count data for Ukraine during 2000-2021 in tiff format. We use tiff formats to illustrate the manipulation of raster data. Such skills will come handy if you ever decide to work with satellite imagery or image data in general.

Before calling the data, let's see how we can use `wpgpDownloadR` package. Let's browse the data catalogue.

```{r}
wpgpListCountries() %>% 
  head()
```

By using the ISO3 country code, let's look for the available datasets for Ukraine.

```{r}
wpgpListCountryDatasets(ISO3 = "UKR") %>% 
  head()
```

The `wpgpDownloadR` package includes 100m resolution data. To keep things efficient, we use 1km gridded population counts from the [WorldPop data page](https://hub.worldpop.org/geodata/listing?id=75). Obtain population data for Ukraine 2000-2020. We start by reading the set of tiff files using the `read_stars` function from the `star` package.

```{r}
#| warning: false

# create a list of file names
file_list <- fs::dir_ls("./data/sequence-analysis/raster")
file_list
# read a list of raster data
pop_raster <- read_stars(file_list, quiet = TRUE)
```

We map the data for 2000 to get a quick understanding of the data.

```{r}
plot(pop_raster[1], col = inferno(100))
```

Next we read shapefile of administrative boundaries in the form of polygons. We obtain these data from the [GADM website](https://gadm.org). GADM provides maps and spatial data for individuals countries at the national and sub-national administrative divisions. In this chapter, we will work with these data as they come directly from the website which provides a more realistic and similar context to which you will probably come across in the "real-world".

```{r}
# read spatial data frame
ukr_shp <- st_read("./data/sequence-analysis/ukr_shp/gadm41_UKR_2.shp") %>% 
  st_simplify(., # simplify boundaries for efficiency
              preserveTopology = T,
              dTolerance = 1000) %>%  # 1km
  sf::st_make_valid(.) %>% 
  fortify(.) %>%  # turns maps into a data frame so they can more easily be plotted with ggplot2
  st_transform(., "EPSG:4326") # set projection system

ukr_shp
```

Let's have a quick look at the resolution of the administrative areas we will be working. The areas below represent areas at the administrative area level 2 in the spatial data frame `ukr_shp`.

```{r}
plot(ukr_shp$geometry)
```

We ensure that the `pop_raster` object is in the same projection system as `ukr_shp`. So we can make both objects to work together.

```{r}
pop_raster <- st_transform(pop_raster, st_crs(ukr_shp))                      
```

### Data wrangling

For our application, we want to work with administrative areas for three reasons. First, public policy and planning decisions are often made based on administrative areas. These are the areas local governments have jurisdiction, represent and can exert power. Second, migration is a key component of population change and hence directly determines population decline. At a small area, residential mobility may also impact patterns of population potentially adding more complexity and variability to the process. Third, WorldPop data are modelled population estimates with potentially high levels of uncertainty or errors in certain locations. Our aim is to mitigate the potential impacts of these errors.

We therefore recommend working with aggregated data. We aggregate the 1km gridded population data to administrative areas in Ukraine. We use `system.time` to time the duration of the proccess of aggregation which could take some time depending on your local computational environment.

```{r}

system.time({

popbyarea_df = aggregate(x = pop_raster, 
                                   by = ukr_shp, 
                                   FUN = sum, 
                                   na.rm = TRUE) 
})

```

**Sub-national population**

The chunk code above returns a list of raster data. We want to create a spatial data frame containing population counts for individual sub-national areas and years. We achieve this by running the following code:

```{r}
# create a function to bind the population data frame to the shapefile
add_population <- function(x) mutate(ukr_shp, 
                      population = x)

# obtain sub-national population counts
ukr_eshp <- lapply(popbyarea_df, add_population)

# create a dataframe with subnational populations
select_pop <- function(x) dplyr::select(x, GID_2, NAME_2, population)
population_df <- lapply(ukr_eshp, select_pop) %>% 
  do.call(rbind, .)
population_df$year <- rep(seq(2000, 2020, by=1), times = 1)
rownames(population_df) <- rep(seq(1, nrow(population_df), by=1), times = 1)

# sub-national spatial data frame
population_df 
```

**National population**

We also create a data frame providing population counts at the national level.

```{r}
# obtain national population counts
population_count <- map_dbl(ukr_eshp, ~.x %>% 
          pull(population) %>% 
          sum(na.rm = TRUE)
        ) %>% 
  as.data.frame()

# change labels
colnames(population_count) <-  c("population")
rownames(population_count) <- rep(seq(2000, 2020, by=1), times = 1)
population_count$year <- rep(seq(2000, 2020, by=1), times = 1)

# national annual population counts
population_count
```

### Exploratory data analysis

Now we are ready to start analysing the data. Before building complexity on our analysis, conducting some exploratory data analysis to understand the data is generally a good starting point, particularly given the multi-layer nature of the data at hand - capturing space, time and population levels. 

**National patterns**

We first analyse national population trends. We want to know to what extent the population of Ukraine has declined over time over the last 20 years. An effective way to do this is to compute summary statistics and visualise the data. Below we look at year-to-year changes in population levels and as a percentage change. By using `patchwork`, we combine two plots into a single figure.

```{r}
# visualise national population trends
pop_level_p <- ggplot(population_count, 
       aes(x = year, y = population/1000000 )) +
  geom_line(size = 1) +
  theme_tufte2() +
  ylim(0, 48) + 
  labs(y = "Population \n(million)",
       x = "Year")
  
# visualise percentage change in population
pop_percent_p <- population_count %>% 
  mutate(
    pct_change = ( ( population - 47955683) / 47955683) * 100
  ) %>% 
ggplot(aes(x = year, y = pct_change )) +
  geom_line(size = 1) +
  theme_tufte2() + 
  labs(y = "Population \npercentage change (%)",
       x = "Year") 

pop_level_p | pop_percent_p
```

**Sub-national**

We also want to know 

```{r}
population_df <- population_df %>% 
  dplyr::group_by(GID_2) %>% 
  arrange(-year, .by_group = TRUE ) %>% 
  mutate(
  pct_change = ( population / lead(population) - 1) * 100, # rate of population change
  pct_change_2000_21 = ( population[year == "2020"] / population[year == "2000"] - 1) * 100, # overall rate of change
  ave_pct_change_2000_21 = mean(pct_change, na.rm = TRUE)
) %>% 
  ungroup()

```

mapping overall change

```{r}

# set colours
cols <- c("#7f3b08", "#b35806", "#e08214", "#faf0e6", "#8073ac", "#542788", "#2d004b")
# reverse order
cols <- rev(cols)

population_df %>% dplyr::filter( year == 2020) %>%
  drop_na(pct_change_2000_21) %>% 
    mutate(
    ove_pop_class = case_when( pct_change_2000_21 <= -3 ~ 'high_decline',
                           pct_change_2000_21 <= -1.5 & pct_change_2000_21 > -3 ~ 'decline',
                           pct_change_2000_21 < -.3 & pct_change_2000_21 > -1.5 ~ 'moderate_decline',
                           pct_change_2000_21 > -0.3 & pct_change_2000_21 < 0.3 ~ 'stable',
                           pct_change_2000_21 > 0.3 & pct_change_2000_21 < 1.5 ~ 'moderate_growth',
                           pct_change_2000_21 >= 1.5 & pct_change_2000_21 < 3 ~ 'growth',
                           pct_change_2000_21 > 3 ~ 'high_growth'),
    ove_pop_class = factor(ove_pop_class, 
         levels = c("high_decline", "decline", "moderate_decline", "stable", "moderate_growth", "growth", "high_growth") )
    ) %>% 
  ggplot(aes(fill = ove_pop_class)) +
  geom_sf(col = "white", size = .1) +
  scale_fill_manual(values = cols,
                    name = "Population change") +
  theme_map_tufte() 
  
 
```

```{r}
population_df %>% dplyr::filter( ave_pct_change_2000_21 < 0) %>% 
  tail(., 40*21) %>% 
  ggplot(data = ., 
           mapping = aes(x= year, y= reorder(NAME_2, pct_change), fill= pct_change)) +
  geom_tile() +
  scale_fill_viridis(name="Population", option ="plasma", begin = .2, end = .8, direction = 1) +
  theme_tufte2() +
  labs(title= paste(" "), x="Year", y="Area") +
  theme(text = element_text(size=14)) + 
  theme(axis.text.y = element_text(size=8))
```

```{r}
# ggplot(data = population_df,
#             mapping = aes(x = year, y = pct_change,
#                           group = NAME_2)) +
#   geom_line()
```

## Application

Explain intuition and main steps

### Defining outcome process

Define the thresholds that should be used to define the extent / pace of acceleration.

```{r}
population_loss_df <- population_df %>% 
  dplyr::filter( pct_change_2000_21 < 0)

population_loss_df %>% 
  dplyr::filter(pct_change  < 0) %>% 
  ggplot(data =  ) +
  geom_density(alpha=0.8, colour="black", fill="lightblue", aes(x = pct_change)) +
  theme_tufte2()
```

Follow the definition of

```{r}
population_loss_df <- population_loss_df %>% 
  dplyr::filter( year != 2000)

population_loss_df <- population_loss_df %>%
  mutate(
    pop_class = case_when( pct_change <= -3 ~ 'high_decline',
                           pct_change <= -1.5 & pct_change > -3 ~ 'decline',
                           pct_change < -.3 & pct_change > -1.5 ~ 'moderate_decline',
                           pct_change > -0.3 & pct_change < 0.3 ~ 'stable',
                           pct_change > 0.3 & pct_change < 1.5 ~ 'moderate_growth',
                           pct_change >= 1.5 & pct_change < 3 ~ 'growth',
                           pct_change > 3 ~ 'high_growth',)
)

```

### Optimal matching

Computation of distance - intuition \> indel operators \> different OM algorithms \> use optimal matching

Rearrange data from long to wide

```{r}
wide_population_loss_df <- population_loss_df %>% 
  as_tibble() %>% 
  group_by(GID_2) %>% 
  arrange(year, .by_group = TRUE ) %>%
  ungroup() %>% 
  tidyr::pivot_wider(
  id_cols =  GID_2,
  names_from = "year",
  values_from = "pop_class"
)
  
wide_population_loss_df
```

```{r}
table(wide_population_loss_df[,2])
```

```{r}
seq.alphab <- c("high_growth", "growth", "moderate_growth", "stable", "moderate_decline", "decline", "high_decline")

seq.lab <- c("High growth", "Growth", "Moderate growth", "Stable", "Moderate decline", "Decline", "High decline")

seq.cl <- seqdef(wide_population_loss_df, 
                 2:21, 
                 alphabet = seq.alphab,
                 labels = seq.lab,
                 cnames = c("2001", "2002", "2003", "2004", "2005", "2006", "2007", "2008", "2009", "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017", "2018", "2019", "2020"),
                 cpal =c("1" = "#7f3b08", 
                         "2" = "#b35806",
                         "3" = "#e08214",
                         "4" = "#faf0e6",
                         "5" = "#8073ac",
                         "6" = "#542788",
                         "7" = "#2d004b"))
```

```{r}
seqplot(seq.cl, 
        title="State distribution plot", 
        type = "d",
        with.legend = "right",
        border = NA)
```

```{r}
# Calculate transition rates
subs_costs <- seqsubm(seq.cl, 
                      method = "TRATE",
                      #time.varying = TRUE
                      )

subs_costs
```

```{r}
seq.trate <- seqtrate(seq.cl)
round(seq.trate, 2)
```

When "auto", the indel is set as `max(sm)/2` when sm is a matrix

```{r}
# Calculate the unique distance matrix
seq.om <- seqdist(seq.cl,
                  method = "OM",
                  indel = "auto",
                  sm = subs_costs)
```

If want to apply varying substitution costs, use `method = DHD`

### Clustering

Different cluster algorithms - use of k-meloids

```{r}
for (k in 2:20)
  pam_sol <- pam(seq.om, k)

```

Determining the number of clusters

```{r}
asw <- numeric(20)
for (k in 2:20)
  asw[k] <- pam(seq.om, k) $ silinfo $ avg.width
  k.best <- which.max(asw)
  cat("silhouette-optimal number of clusters:", k.best, "\n")
  asw
```

```{r}

pam_optimal <- pam(seq.om, 7)

```

### Visualising

Individual trajectories

```{r}
seqplot(seq.cl, 
        group = pam_optimal$clustering,
        type = "I",
        border = NA, 
        cex.axis = 1.5, 
        cex.lab = 1.5,
        sortv = seq.om)
```

```{r}
ggseqiplot(seq.cl, 
        group = pam_optimal$clustering,
        sortv = seq.om,
        facet_ncol = 4) +
  scale_fill_manual(values = rev(cols)) +
  scale_color_manual(values = rev(cols)) 
```

Frequency plots

```{r}
ggseqdplot(seq.cl, 
        group = pam_optimal$clustering,
        facet_ncol = 4) +
  scale_fill_manual(values = rev(cols)) +
  scale_color_manual(values = rev(cols)) 
```

Time spent

```{r}
ggseqmtplot(seq.cl, 
        group = pam_optimal$clustering,
        facet_ncol = 4) +
  scale_fill_manual(values = rev(cols)) +
  scale_color_manual(values = rev(cols)) +
  scale_x_discrete(labels=c("high_growth" = "HG", "growth" = "G",
                              "moderate_growth" = "MG", "stable" = "S", "moderate_decline" = "MD", "decline" = "D", "high_decline" = "HD" ))
```

Entropy index

```{r}
ggseqeplot(seq.cl, 
        group = pam_optimal$clustering) +
  scale_colour_viridis_d()
```

## Questions

Francisco to write

This chapter shows the use of sequence analysis to study population decline and draw on the following papers:

-   @rowe2022a

-   @newsham2022a

-   @rowe2022b
